Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-11-15 ms/ms.Rnw
final biorxiv version

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Twitter handle:} @thegoodphyte \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 199\\
Introduction: 1355\\
Materials and Methods: 1698\\
Results: 582\\
Discussion: 1935\\
Acknowledgement: 26\\
4 Figures and 1 Table, 5 Supplemental Figures

@@

%\linenumbers
%\setstretch{\stretchy}
\pagenumbering{gobble}
%\raggedright

@@

Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r life form, stomata, stomatal ratio

@@

% REMOVED: I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface.

@@

% REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

@@

The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}. It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of palisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface. Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous. Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands \citep{Richardson_etal_2017}. Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. The cost of adaxial stomata may be greater in the shade because wetter leaves and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.

@@

I refer to the second hypothesis as the `coordination' hypothesis. In this scenario, ecological conditions such as high light select for both increased total stomatal density and for amphistomy because these traits work well in coordination with one another. For example, if stomatal density were very high on a hypostomatous plant, then CO$_2$ would be more strongly limited by the mesophyll. Adding a second parallel pathway for diffusion by developing stomata on both surfaces would restore a more optimal balance between stomatal and mesophyll limitations. Conversely, there would be little benefit to amphistomy when total stomatal density is low because CO$_2$ diffusion is strongly limited by stomatal resistance, and therefore photosynthetic rate is not sensitive to changes in mesophyll diffusion mediated by stomatal ratio. A related prediction is that increased atmospheric CO$_2$ may select for reduced stomatal ratio and density primarily by decreasing adaxial stomatal density, but this has not been well tested \citep[but see][]{Woodward_Bazzaz_1988}. These results suggest that coordination between stomatal ratio and density might play a greater role than previously appreciated in optimizing CO$_2$ supply and demand under different light regimes \citep[see also][]{Beerling_Kelly_1996}.

@@

\section*{Supporting figure captions}

Figure S1: Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface

Figure S2: Raunki\ae r life form and plant habit broadly overlap.

Figure S3: Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.

Figure S4: Growth forms have different tolerances for sun and shade among British angiosperms.

Figure S5: The effect of light on stomatal ratio depends on growth form.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-lf.pdf}}
%\centerline{\includegraphics[angle=270]{figures/figure_SR-lf.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity. Key to symbols: n.s. = not significant,  $^*~P < 0.05$, $^{**}~P < 0.01$, $^{***}~P < 0.001$.} 
\label{fig:SR-lf}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-hf.pdf}}
%\centerline{\includegraphics[angle=270]{figures/figure_SR-hf.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on growth form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in annual, biennual, and perennial herbs, but not shrubs or trees. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to growth form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity. Key to symbols: n.s. = not significant,  $^*~P < 0.05$, $^{**}~P < 0.01$, $^{***}~P < 0.001$.} 
\label{fig:SR-hf}
\end{figure}
\end{document}

>>>>

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\begin{center}
In press at \textit{New Phytologist} \\
\end{center}
\clearpage
%\textit{Twitter handle:} @thegoodphyte \\
%\\
%\textit{Author for correspondence:} \\
%\textit{Christopher D. Muir} \\
%\textit{Tel:} +1 984-218-5309 \\
%\textit{Email:} chrisdmuir@gmail.com \\
%1413 N Mangum St\\
%Durham, NC 27701\\
%\\
%Short title: Shedding light on stomatal evolution\\
%\\
%Word count: \\
%Summary: 199\\
%Introduction: 1355\\
%Materials and Methods: 1698\\
%Results: 582\\
%Discussion: 1935\\
%Acknowledgement: 26\\
%4 Figures and 1 Table, 5 Supplemental Figures

@@

%\linenumbers
%\setstretch{\stretchy}
%\pagenumbering{gobble}
%\raggedright

@@

adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r life form, stomata, stomatal ratio

@@

@@

@@

The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}. Additional CO$_2$ should increase quantum yield under high light when chloroplastic CO$_2$ concentration strongly limits photosynthetic rate. The association between light tolerance and amphistomy is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of palisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface. Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous. Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands \citep{Richardson_etal_2017}. Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. The cost of adaxial stomata may be greater in the shade because wetter leaves and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.

@@

I refer to the second hypothesis as the `coordination' hypothesis. In this scenario, ecological conditions such as high light select for both increased total stomatal density and for amphistomy because these traits work well in coordination with one another. For example, if stomatal density were very high on a hypostomatous plant, then CO$_2$ would be more strongly limited by the mesophyll \citep{Flexas_etal_2012}. Adding a second parallel pathway for diffusion by developing stomata on both surfaces would restore a more optimal balance between stomatal and mesophyll limitations. Conversely, there would be little benefit to amphistomy when total stomatal density is low because CO$_2$ diffusion is strongly limited by stomatal resistance, and therefore photosynthetic rate is not sensitive to changes in mesophyll diffusion mediated by stomatal ratio. A related prediction is that increased atmospheric CO$_2$ may select for reduced stomatal ratio and density primarily by decreasing adaxial stomatal density, but this has not been well tested \citep[but see][]{Woodward_Bazzaz_1988}. These results suggest that coordination between stomatal ratio and density might play a greater role than previously appreciated in optimizing CO$_2$ supply and demand under different light regimes \citep[see also][]{Beerling_Kelly_1996}.

@@

%\section*{Supporting figure captions}

%Figure S1: Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface

%Figure S2: Raunki\ae r life form and plant habit broadly overlap.

%Figure S3: Phylogenetic diversification of stomatal ratio follows growth form and light tolerance.

%Figure S4: Growth forms have different tolerances for sun and shade among British angiosperms.

%Figure S5: The effect of light on stomatal ratio depends on growth form.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-lf.pdf}}
%\centerline{\includegraphics[angle=270]{figures_eps/figure_SR-lf.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity. Key to symbols: n.s. = not significant,  $^*~P < 0.05$, $^{**}~P < 0.01$, $^{***}~P < 0.001$.} 
\label{fig:SR-lf}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-hf.pdf}}
%\centerline{\includegraphics[angle=270]{figures_eps/figure_SR-hf.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on growth form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in annual, biennual, and perennial herbs, but not shrubs or trees. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to growth form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity. Key to symbols: n.s. = not significant,  $^*~P < 0.05$, $^{**}~P < 0.01$, $^{***}~P < 0.001$.} 
\label{fig:SR-hf}
\end{figure}
\end{document}

### 2017-11-15 ms/ms.tex
final biorxiv version

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Twitter handle:} @thegoodphyte \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 199\\
Introduction: 1355\\
Materials and Methods: 1698\\
Results: 582\\
Discussion: 1935\\
Acknowledgement: 26\\
4 Figures and 1 Table, 5 Supplemental Figures

%\linenumbers
%\setstretch{\stretchy}
\pagenumbering{gobble}
%\raggedright

>>>>

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\begin{center}
In press at \textit{New Phytologist} \\
\end{center}
\clearpage
%\textit{Twitter handle:} @thegoodphyte \\
%\\
%\textit{Author for correspondence:} \\
%\textit{Christopher D. Muir} \\
%\textit{Tel:} +1 984-218-5309 \\
%\textit{Email:} chrisdmuir@gmail.com \\
%1413 N Mangum St\\
%Durham, NC 27701\\
%\\
%Short title: Shedding light on stomatal evolution\\
%\\
%Word count: \\
%Summary: 199\\
%Introduction: 1355\\
%Materials and Methods: 1698\\
%Results: 582\\
%Discussion: 1935\\
%Acknowledgement: 26\\
%4 Figures and 1 Table, 5 Supplemental Figures

%\linenumbers
%\setstretch{\stretchy}
%\pagenumbering{gobble}
%\raggedright
